Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2024-11-19 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{figure}[]
    \centering
    \vspace{-.25cm}
    \includegraphics[width=1\linewidth]{figures/sbi.png}
    \vspace{-.7cm}
    \caption{SBI flow chart. From a set of priors we simulate a sample of stellar abundances using \texttt{CHEMPY} which we use to train a Neural Density Estimator. With this we infer the posterior distribution of the model parameters from a single star. Repeating that for $N_{\rm stars}$ from the same galaxy gives an accurate fit of the IMF slope and Type Ia supernovae normalization.}
    \label{fig:flowchart}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/sbi_Nstar_comp.png}
    \vspace{-.25cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue) to the inferred values using HMC (red) \cite{Philcox_2019} .}
    \label{fig:N_star_analysis}
\end{figure}

@@

\paragraph{Inference}
\begin{figure}
\vspace{-1.3cm}
    \centering
    \includegraphics[width=0.49\textwidth]{figures/posterior_2.png}
    \vspace{-.25cm}
    \caption{Sampled posteriors for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars.}
    \label{fig:sbi}   
\end{figure}
\enlargethispage{0\baselineskip}
For each set of mock abundances we infer the posterior distribution for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with the NPE by sampling 1000 times and fitting the posterior for each parameter. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. 
Our method takes around $5$ minutes to build a posterior for all six parameters for a dataset size of 1000 stars on a Macbook Air M2 chip, making it around $2400$x faster than current HMC methods \citep[cf.][who needs $40$h for 200 stars with HMC]{Philcox_2019}. Hence, shorter computing times make it feasible to use orders of magnitudes more observations.
%
Taking the means of multiple observed stars gives us higher accuracy and precision of the global galactic parameters $\vec\Lambda$ (Fig. \ref{fig:N_star_analysis}). However, we also see that in the limit of very few stars (less than $\sim10$) SBI shows a larger bias than HMC results. But in the limit of large numbers of stars (few hundred) the accuracy and precision is superior compared to HMC.
%
To infer the global galactic parameters $\vec\Lambda$ of the stars as accurate as possible we take the mean of all inferred posteriors from $n_{\rm stars}=1000$ (Fig. \ref{fig:sbi}) resulting in $\alpha_{\rm IMF}=-2.292\pm0.003$ and $\log_{10}(\rm N_{Ia})=-2.883\pm0.005$ which deviates less than $0.4\%$ from the correct value.

>>>>

% \begin{figure}[]
%     \centering
%     \vspace{-.25cm}
%     \includegraphics[width=1\linewidth]{figures/sbi.png}
%     \vspace{-.7cm}
%     \caption{SBI flow chart. From a set of priors we simulate a sample of stellar abundances using \texttt{CHEMPY} which we use to train a Neural Density Estimator. With this we infer the posterior distribution of the model parameters from a single star. Repeating that for $N_{\rm stars}$ from the same galaxy gives an accurate fit of the IMF slope and Type Ia supernovae normalization.}
%     \label{fig:flowchart}
% \end{figure}

@@

% \begin{figure}
%     \centering
%     \includegraphics[width=0.9\linewidth]{figures/sbi_Nstar_comp.png}
%     \vspace{-.25cm}
%     \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue) to the inferred values using HMC (red) \cite{Philcox_2019} .}
%     \label{fig:N_star_analysis}
% \end{figure}

@@

\paragraph{Inference}

% \begin{figure}
% \vspace{-1.3cm}
%     \centering
%     \includegraphics[width=0.49\textwidth]{figures/posterior_2.png}
%     \vspace{-.25cm}
%     \caption{Sampled posteriors for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1000$ stars.}
%     \label{fig:sbi}   
% \end{figure}

For each set of mock abundances we infer the posterior distribution for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with the NPE by sampling 1000 times and fitting the posterior for each parameter. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. 
Our method takes around $5$ minutes to build a posterior for all six parameters for a dataset size of 1000 stars on a Macbook Air M2 chip, making it around $2400$x faster than current HMC methods \citep[cf.][who needs $40$h for 200 stars with HMC]{Philcox_2019}. Hence, shorter computing times make it feasible to use orders of magnitudes more observations.
%
Taking the means of multiple observed stars gives us higher accuracy and precision of the global galactic parameters $\vec\Lambda$ (Fig. \ref{fig:N_star_analysis}). However, we also see that in the limit of very few stars (less than $\sim10$) SBI shows a larger bias than HMC results. But in the limit of large numbers of stars (few hundred) the accuracy and precision is superior compared to HMC.
%
To infer the global galactic parameters $\vec\Lambda$ of the stars as accurate as possible we take the mean of all inferred posteriors from $n_{\rm stars}=1000$ (Fig. \ref{fig:sbi}) resulting in $\alpha_{\rm IMF}=-2.292\pm0.003$ and $\log_{10}(\rm N_{Ia})=-2.883\pm0.005$ which deviates less than $0.4\%$ from the correct value.
